---
title: "HOW CAN A WELLNESS COMPANY CAN PLAY IT SMART"
author: "OLABISI FAYOSE"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: cerulean
    toc: true
--- 

## CLIENT : BELLABEAT

## EXECUTIVE SUMMARY

* Fitness trackers are electronics technological devices incorporated into items that can be comfortably worn on a body for monitoring fitness related metrics.

* The wearable  fitness trackers market was valued at  in $17.9M in 2016,  $45B in 2021 and projected to reach $192B in 2030 at a CAGR of 17.5%. The expansion is due to Increasing Health awareness among the consumers and the value of living a healthy lifestyle.

* FITBIT is a consumer electronics and fitness company. It produces wireless enabled wearable fitness monitors. In 2016, FitBit was the market leader with a market share of 22%. BellaBeat on the other hand is a new entrant in the wearable fitness monitors market in a unique category for women.

* **THE  PROBLEM**

Although BellaBeat is quite successful but,they want to be a  major player in the Global market. They are interested in knowing more about consumer trends and key  marketing success factors in the fitness tracker market in order to help develop their marketing strategy.

## PROJECT OBJECTIVE/BUSINESS TASK
* FOCUS ON ONE BELLABEAT PRODUCT
* ANALYZE A  SMART DEVICE DATA TO GAIN INSIGHT INTO HOW CONSUMERS ARE USING THEIR SMART DEVICE. 
* USE THE INSIGHTS GENERATED TO GUIDE A MARKETING STRATEGY FOR THE BELLABEAT COMPANY

## The Dataset
* A kAGGLE open source Dataset consisting of  18 Tables
Generated by  30 FitBit users Respondents to a distributed survey via Amazon mechanical turk between 03.12.2016 and 05.12.2016
* 30 eligible FitBit users concentrated to the submission of personal trackers data including minute by minute output for physical activities, heart rate, sleep monitoring, calories burnt and METs.
* Variations in output represents use of different types of FitBit trackers.

### DATA INTEGRITY
* using the ROCCC test, the dataset is certified credible for the data analysis
 
reliable
Original,
Comprehensive - it shows how consumers use their trackers
Current - based on assumption the analysis demand was made in 2017
Cited - cited
* It is also unbiased.

### CLEANING
* NO DUPLICATES found
* No null values found
* Convertion of dates from character class to date class was carried out where necessary. Details are in the findings

## Market trends
* Global fitness tracker market size went from USD17 million in 2016 to $41 billion in 2021 and is expected to be 114B by year 2028 at a CAGR of 15.4%.[link](www.fortunebusinessinsights.com) 
* The growth is attributed to increase in healthcare awareness, rise in the number of manufacturers offering various fitness trackers at affordable prices and increased prevalence of Health disorders like obesity, diabetes, cardiovascular diseases etc.Another of the major causes of the surge was COVID_19 pandemic.
*Growth limited by risk of data theft

###MARKET SHARE

* Market share as illustrated in the chart. Market dominated by Apple.

```{r}
library(tidyverse)
```


```{r}
mark <- read.csv("C://Users//ACER//Downloads//Untitled spreadsheet - Sheet1 (1).csv")
```
```{r}
head(mark)
```
```{r}
mark %>% 
  ggplot(aes(YEAR, MARKET.SHARE, color = BRAND))+
  geom_line(size = 2)+
  labs(title = "MARKET SHARES TRENDS FROM 2014 - 2021")
```


### MARKET SEGMENTS

#### DEVICE TYPE SEGMENT
* Smartwatches  
* WristBands
* Smart Glasses   
* Smart Clothings
* Others
 Dominated by SmartWatches 40% in 2020

#### APPLICATION SEGMENT
* Running                 
* Heart Rate
* Cycling                  
* Sleep
* Sports                      
* Glucose 
 Dominated by Running 
 #### DISTRIBUTION CHANNEL SEGMENT
* Online
* Offline
 Dominated by Online61% in 2020

#### REGIONAL SEGMENT
* North America  
* Europe
* Asia Pacific            
* Latin America
* Middle East & Africa
Dominated by North America 42% in 2020

## FINDINGS
### GENERAL 
* There were different number of respondents tracked the different applications. Maybe because they had different devices with different functions
* Not all respondents track everyday 
* Applications tracked include
  1. Physical Activities 2. Steps 3. Heart rate, 4. Sleep, 5. Calories burnt, 6. METs
* IMPORT FILES. Although there are 18 data sets, we will not be importing all because some of them have been merged into one dataframe viz: Activities, steps, calories burnt, time spent on activities etc.
* Set up the environment by running these libraries

```{r}
library(tidyverse)
```
```{r}
library(GGally)
```
```{r}
library(lubridate)
```
```{r}
library(ggplot2)
```

### sleep

```{r}
sleep <- read.csv("C://Users//ACER//Downloads//sleepDay_merged.csv")
```

* Only 24 respondents tracked their sleep. 

```{r}
unique(sleep$k)
```

* time spent in bed was tracked and differentiated from time spent sleeping
*time spent sleeping

* Recommended sleep time is 7-9hours a day. Average sleep time is 6.9hours. See code below.
```{r}
summary(sleep$TOTAL.HOURS)
```

* quality of sleep was tracked and segregated into 3 categories 1, 2 3. 1== Deep sleep, 2 == REM or RAPID EYE MOVEMENT, 3 == light sleep
Of which deep sleep is the best.Respondents generally sleep deeply as shown below.
In the dataset quality of sleep was recorded as 1, 2, 3, but i later did a recode to deep, REM and light for better understanding. See code chunck below
```{r}
sleep$TotalSleepRecords = factor(sleep$TotalSleepRecords)
sleep$TotalSleepRecords2[sleep$TotalSleepRecords == "1"] = "deep"
sleep$TotalSleepRecords2[sleep$TotalSleepRecords == "2"]  = "rem"
sleep$TotalSleepRecords2[sleep$TotalSleepRecords == "3"] = "light" 
```

```{r}
sleep %>% 
  ggplot(aes(TOTAL.HOURS.1,fill = TotalSleepRecords2))+
  geom_histogram(binwidth = 1)+
  labs(title = "DISTRIBUTION OF SLEEP QUALITY")
```


* Respondents sleep for longer hours on Saturdays.See corresponding charts below
```{r}
library(lubridate)

```
First we have to change the format of the date in the sleep dataset from character to date using the code below.
```{r}
class(sleep$SleepDay)
```

```{r}
sleep$DAT <- parse_date_time(sleep$SleepDay, "%a %B %d,%Y")
```
```{r}
class(sleep$DAT)
```
```{r}
head(sleep$DAT)
```

Now that a new date colum has been created from sleepdate column, day of the week will be extracted to form another column
```{r}
sleep$wday <- wday(sleep$DAT, label = TRUE)
```
```{r}
head(sleep$wday)
```

```{r}
summary(sleep$TOTAL.HOURS)
```
A plot of sleep time on different days of the week is below
```{r}
sleep %>% 
  ggplot(aes(sleep$TOTAL.HOURS, color = sleep$wday))+
  geom_density()+
  labs(title = "SLEEPTIME ON DIFFERENT WEEKDAYS")

  
```

``

### DAILY ACTIVITIES AND CALORIES BURNT

* Physical activies, number of steps, time spent in different activies and calories burnt  were tracked differently per minute but their results were later merged in a single dataset making analysis easier.
Dataset saved as daily below.`
```{r}
daily <- read.csv("C://Users//ACER//Downloads//dailyActivity_merged (1).csv")
```
```{r}
summary(daily)
```
```{r}
daily %>% 
  select(TotalSteps,VeryActiveDistance,SedentaryMinutes,Calories) %>% 
  filter(TotalSteps > 0) %>% 
  ggplot(aes(x =TotalSteps, y = Calories))+
  geom_point(size = 1.5,alpha = 1, color = "red")+
  geom_smooth(method = "lm")+
  theme_bw()+
  labs(title = "CORRELATION BETWEEN CALORIES AND TOTALSTEPS")
```
* comparing rate of activities per day, respondents are generally non active. They live more of a sedentary lifestyle.
```{r}
time <- c(1440, 900, 192, 13.6, 21.6)
```
```{r}
active <- c("totalday", "sedentary", "light", "moderate", "veryactive")
```
```{r}
activetime <- data.frame(time, active)
```
```{r}

```


```{r}
bp<- ggplot(activetime, aes(x="", y= time, fill = active))+
  geom_bar(width = 1, stat = "identity")
```
```{r}
pie <- bp + coord_polar("y", start=0)+
  labs(title = "AVERAGE TIME SPENT ON DIFFERENT ACTIVITIES")
pie
```

* Carrying out a correlation between the different variables of daily activities versus calories. I first removed non numeric and  irrelevant variables with the code beneath 

```{r}
dailycor <- subset(daily, select = -c(Id,ActivityDate,LoggedActivitiesDistance,SedentaryActiveDistance,TotalDistance ))
```
```{r}
names(dailycor)
```

* secondally i installed the GGally package and made the plots as shown below
l
```{r}
library(GGally)
```
```{r}
ggpairs(dailycor)
```
```{r}
ggcorr(dailycor, name = expression(rho), geom = "circle", max_size = 10, min_size = 2, size = 3, hjust = 0.75, 
       nbreaks = 6, angle =-90)
```

* from the correlation plots above and more detailed beneath, calories burnt per day are correlated to number of steps, active distance covered per day and very active minutes.

```{r}
daily %>% 
  select(TotalSteps,TrackerDistance,VeryActiveDistance,SedentaryMinutes,Calories) %>% 
  filter(TotalSteps > 0) %>% 
  ggplot(aes(x =VeryActiveDistance, y = Calories ))+
  geom_point()+
  geom_smooth(method = "lm")+
  labs(title = "CORRELATION BETWEEN CALORIES BURNT AND VERY ACTIVEDISTANCE")


```
* Calories burnt per day is also correlated with TotalSteps. see plot below.

```{r}
daily %>% 
  select(TotalSteps,VeryActiveDistance,SedentaryMinutes,Calories) %>% 
  filter(TotalSteps > 0) %>% 
  ggplot(aes(x =TotalSteps, y = Calories))+
  geom_point(size = 1.5,alpha = 1, color = "red")+
  geom_smooth(method = "lm")+
  theme_bw()+
  labs(title = "CORRELATION BETWEEN CALORIES BURNT AND TOTALSTEPS")

```


* below are distribution plots of some variables e.g. Total Steps and tracked distance
```{r}
daily %>% 
  select(TotalSteps,TrackerDistance,VeryActiveDistance,SedentaryMinutes,Calories) %>% 
  filter(TotalSteps > 0) %>% 
  ggplot(aes(TotalSteps))+
  geom_histogram(bins = 30, color = "red")+
  labs(title = "DISTRIBUTION OF TOTALSTEPS")

```
```{r}
daily %>% 
  select(TotalSteps,TrackerDistance,VeryActiveDistance,SedentaryMinutes,Calories) %>% 
  filter(TotalSteps > 0) %>% 
  ggplot(aes(Calories))+
  geom_histogram(bins = 30, color = "blue")+
  labs(title = "DISTRIBUTION OF VERY ACTIVE DISTANCE")
```



### HEARTRATE

* 14 participants
```{r}
heart<- read.csv("C://Users//ACER//Downloads//heartrate_seconds_merged.csv (1)//heartrate_seconds_merged.csv") 
```
```{r}
unique(heart$Id)
```


 * Recorded every seconds. Making it impossible to combine with other datasets which are in minutes.
 Below is a chart of each individual heart beat rate distribution.
```{r}
heart$Id  = factor(heart$Id)
```
```{r}
heart %>% 
  ggplot(aes(Value, color = Id))+
  geom_boxplot()+
  labs(title = "THE DIFFERENT RESPONDENTS AND THEIR HEARBEATS RATE")

```
 
 ### METs
 * MET is an abbreviation of Metabolic equivalents and it measures the body's expenditure of energy. 
* The higher the met, the more rigorous is the activity you are engaging
* MET is also function of calories burnt using the formula below. The higher your MET, the higher is the calories burnt.
 ***(MET * 3.5 * body weight)/200 = Calories burnt per minute***
 
* 33 respondents
```{r}
mets <- read.csv("C://Users//ACER//Downloads//minuteMETsNarrow_merged.csv//minuteMETsNarrow_merged.csv")

```

```{r}
unique(mets$Id)
```


 * Below is the summary of the dataset
 
```{r}
summary(mets$Id)
```
 
* Below is a distribution of the METs values. It shows a largely sedentary lifestyle with few exceptions exerting more than average energy expenditure and engaging in rigorous activities.

```{r}
mets %>% 
  ggplot(aes(METs))+
  geom_density(color = "red")+
  labs(title = "DISTRIBUTION OF METs/MINUTE")

```

## BELLABEAT LEAF

* Wristband, brooch or a necklace Fitness tracker

### STRENGHTS

*Aesthetically attractive. Looks like a piece of Jewelry

* Uniquely positioned for women
* Tracks activities, sleep, steps, stress, reproductive health
* long battery life (six months). no need to keep recharging weekly
* Medium price
* for sale online from company website and some major online stores

### WEAKNESSES
* Positioned mainly for women
* Limited application tracking
* Need to change position of leaf for monitoring different activities
* Not available in physical stores. 

## CONCLUSION/RECOMMENDATION

* The fitness trackers market is highly dynamic and constantly changing as new devices are released, relaunching of existing brands, withdrawal of non performing brands, new vendors entering or leaving or acquired by bigger vendors.How do you explain FITBIT going from biggest market share holder in 2016 to insignificant share today.
* The market is innovation technology driven.
* This dataset showed how consumers use their trackers and the different applications used at that time 2016-2017. Many Consumers dont use their trackers everyday.
* BellaBeat trackers don't track enough applications compared to FITBIT at the this assumed epoch 2016-2017. BellaBeat needs to increase applications of their trackers to include for instance heart beat rate, calories burnt, glucose etc and be more proactive.
 
* BellaBeats need to increase application types including heartbeat, glucose etc.

* Why limit BellaBeat products to women only. Why not men or unisex products as well. Limiting products to women only reduces target audience, product purchase and shares by almost half.

* if 60% consumers purchase trackers online, there is a need to increase online presence especially advertisement. There is need to sell in physical stores to cater for 40% of target audience that buy offline.

* North America controls almost half of buyers followed by Europe. However, Asia seems to be an emerging market with  great potentials for growth due to their large population. Distribution plus advertisement of BellaBeat products in these areas should be emphasized.
* Finaly, there is need for further research to compare  performance of BellaBeat Products with key brands.It will also serve as a feedback from consumers on evaluation of the BellaBeat products, their strengths as well as areas of improvement.